---
title: "finalproject"
output:
  html_document: default
  pdf_document: default
date: "2022-11-09"
---


```{r setup}

#I am introducing the package "geomorph", which is used 
#for statistical analyses of 2D and 3D geometric
#morphometric landmark data
library(geomorph)

#I am also using "ggtree", "ape" and "knitr" for graphic
#visualization and phylogenetic analysis
library(ggtree)
library(ape)
library(knitr)

#setting working directory
setwd("C:/Users/18034/Dropbox/Senior Year/Phylogenetic Biology/Final Project")

```

```{r}

#identifying file path
file.path("C:\\Users\\18034\\Dropbox\\Senior Year\\Phylogenetic Biology\\Final Project\\simulation.tps")

#reading and plotting the tps file, while contains 
#the 3D landmark data

#for this analysis, I simulated the number of specimens
#I anticipate having for my final analysis. I plotted 
#the landmarks using tipsUtil64 and tipsdig264, and
#arbitrairily marked 5 landmark points per specimen.

SimulationData <- readland.tps("simulation.tps", specID = "ID")

#This plot displays all landmark points on one plot,
#uncorrected for orientation or scale.

plot(SimulationData)

```
```{r}

#this function aligns and scales the landmark 
#data for all specimens. It uses Procrustes 
#analysis to rescale the landmarks for 
#different sizes of bones and reorients the 
#shapes around the origin.

Simulationgpa <- gpagen(SimulationData)

#this function plots the new data that has
#been adjusted with Procrustes analysis.

plot(Simulationgpa)

```

``` {r}

#This function introduces an established tetrapod 
#tree for comparison against landmark data (Pyron & 
#Jetz 2018). It includes all tetrapods, and will
#need to be pruned.

jetz.tree = read.tree("amph_shl_new.tre")

#This function creates a vector to isolate the tips
#of interest. In this case, that includes all
#Plethodontidae species that are both included on 
#the tree and in geometric morphometric analysis.

plethodontidae <- c("Aneides_aeneus","Aneides_hardii","Aneides_lugubris","Batrachoseps_pacificus","Bolitoglossa_lincolni", "Bolitoglossa_mexicana", "Desmognathus_fuscus", "Ensatina_eschscholtzii", "Eurycea_lucifuga", "Eurycea_quadridigitata", "Hemidactylium_scutatum", "Hydromantes_platycephalus", "Plethodon_cheoah", "Plethodon_cinereus", "Plethodon_dunni", "Plethodon_elongatus", "Plethodon_glutinosus", "Thorius_boreas")

print(plethodontidae)

#This function prunes the tetrapod tree, keeping 
#only the tips of interest. That includes all 
#species within our defined "plethodontidae" vector.

pr.jetztree <- keep.tip(jetz.tree, plethodontidae)

#This function plots the newly pruned tree.

plot(pr.jetztree)


```


```{r}

#visualizing feeding types

#Here, the data.frame function is used to create
#a dataset of species of interest and their 
#respective feeding mechanisms (muscular 
#(indicated by "m") or ballistic (indicated by 
#"b")).

feeding.behavior = data.frame(
  species = c("Aneides_aeneus","Aneides_hardii","Aneides_lugubris","Batrachoseps_pacificus","Bolitoglossa_lincolni", "Bolitoglossa_mexicana", "Desmognathus_fuscus", "Ensatina_eschscholtzii", "Eurycea_lucifuga", "Eurycea_quadridigitata", "Hemidactylium_scutatum", "Hydromantes_platycephalus", "Plethodon_cheoah", "Plethodon_cinereus", "Plethodon_dunni", "Plethodon_elongatus", "Plethodon_glutinosus", "Thorius_boreas"),
  feeding.type = c("m","m","m","b","b","b","m","b","b","b","b","b","m","m","m","m","m","b"))


#This function plots the pruned tree with the 
#data set "feeding.behavior", mapping character 
#states onto the tip nodes using known information.

feeding.plot = ggtree( pr.jetztree ) %<+% feeding.behavior + 
  geom_tiplab( fontface = "italic", offset=0.02 ) +
  xlim(0, 0.5) +
  geom_tippoint( aes(colour=feeding.type), size=3, alpha=5 )

feeding.plot


```
```{r}
#plotting scalar landmark data

#As above, the data.frame function creates
#a data set of species of interest and a
#trait of interest. In this case, it is 
#scapula length, a continuous variable. The
#specific scapula lengths have yet to be 
#extracted from the landmark data, so I
#am currently using made-up numbers.

scalar.scapula = data.frame(
  species = c("Aneides_aeneus","Aneides_hardii","Aneides_lugubris","Batrachoseps_pacificus","Bolitoglossa_lincolni", "Bolitoglossa_mexicana", "Desmognathus_fuscus", "Ensatina_eschscholtzii", "Eurycea_lucifuga", "Eurycea_quadridigitata", "Hemidactylium_scutatum", "Hydromantes_platycephalus", "Plethodon_cheoah", "Plethodon_cinereus", "Plethodon_dunni", "Plethodon_elongatus", "Plethodon_glutinosus", "Thorius_boreas"),
  length.value = c(1,2,3,4,5,4,3,2,1,2,3,4,5,6,5,4,3,2))


#This function plots the pruned tree with the 
#data set "scalar.scapula", mapping character 
#states onto the tip nodes using known information.

scalar.plot = ggtree( pr.jetztree ) %<+% scalar.scapula + 
  geom_tiplab( fontface = "italic", offset=0.02 ) +
  xlim(0, 0.5) +
  geom_tippoint( aes(colour=length.value), size=3, alpha=5 )

scalar.plot

```

```{r}

#This function creates a data frame of 
#the landmark data, with reference to 
#the Jetz phylogeny. This is a 
#requirement of the geomorph package 
#for running PGLS analyses.

gdf <- geomorph.data.frame(Simulationgpa, phy = pr.jetztree)
attributes(gdf)

#This function performs PGLS with the 
#established data frame, which includes 
#landmark data and the pruned phylogeny.

simulationPGLS <- procD.pgls(coords ~ Csize, phy=pr.jetztree, iter=999, data=gdf)
anova(simulationPGLS)
summary(simulationPGLS)
```




```{r}
#note to henry: i want to add some sort of ggplot that puts the feeding behavior (like in the previous plot) against a gradient of scapula shape. im thinking i can probably do it with ggplot, and maybe ill just convert one of the landmark pairs to a scalar quality and plot that? would love to hear your thoughts! I also dont know if im running the pgls right, like if im actually including variables of interest, so pls lmk